---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r base R version}
version 
```

As of 2022-11-28:
version.string R version 4.2.2 (2022-10-31 ucrt)
nickname       Innocent and Trusting  




```{r tidyverse versions}
library(tidyverse)

sessioninfo::session_info() %>%
  pluck(2) %>%
  as_tibble() %>%
  filter(attached == TRUE) %>%
  select(package, loadedversion) %>%
  knitr::kable()
```

As of 2022-11-28:

|package   |loadedversion |
|:---------|:-------------|
|dplyr     |1.0.10        |
|forcats   |0.5.2         |
|ggplot2   |3.4.0         |
|purrr     |0.3.5         |
|readr     |2.1.3         |
|stringr   |1.4.1         |
|tibble    |3.1.8         |
|tidyr     |1.2.1         |
|tidyverse |1.3.2         |






Note on native pipe `|>` vs the magrittr `%>%` pipe: 

While the native pipe is more efficient operationally, the magrittr pipe has some more flexible functionality.  

Due to the small nature of these data sets, the increased efficiency of the native pipe is not necessary and the flexibility of the magrittr pipe is preferred. 
